---
title: "Bowhead whale track analysis with LangevinAnimalTracking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bowhead whale track analysis with LangevinAnimalTracking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

This vignette demonstrates how to fit the penalized Langevin movement model to
a real bowhead whale telemetry track using the SGD estimation algorithm, and
how to compare the results with the smoothSDE Kalman-filter baseline.

The data consist of Argos satellite telemetry locations for multiple individuals
tagged in Hudson Bay / Foxe Basin (Canada). We focus on a single individual
(ID 128145) for illustration.

## 1. Setup and data loading

```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(sf)
library(sp)
library(rnaturalearth)
library(rnaturalearthhires)
library(ggplot2)
library(smoothSDE)
library(LangevinAnimalTracking)

set.seed(42)
```

```{r load-data}
data_path <- system.file("extdata", "bowhead_data.csv",
                         package = "LangevinAnimalTracking")
data <- read.csv(data_path) |>
  mutate(
    timestamp = ymd_hms(timestamp),
    whale_id  = as.factor(individual.local.identifier)
  )

# Keep locations inside the study area
bbox_lims <- c(bottom = 61, top = 72.5, left = -95, right = -70)

data <- data |>
  filter(
    location.lat  >= bbox_lims["bottom"],
    location.lat  <= bbox_lims["top"],
    location.long >= bbox_lims["left"],
    location.long <= bbox_lims["right"]
  )
```

## 2. Exploratory visualisation

```{r land-polygon}
bbox      <- st_bbox(c(xmin = -95, xmax = -70, ymin = 61, ymax = 72.5),
                     crs = st_crs(4326))
bbox_poly <- st_as_sfc(bbox)

land <- ne_countries(scale = "large", returnclass = "sf")
land <- st_make_valid(land)
land <- st_crop(land, bbox)
```

```{r plot-all-individuals}
ggplot() +
  geom_sf(data = land, fill = "grey85", colour = "grey60",
          linewidth = 0.25) +
  geom_path(data = data,
            aes(location.long, location.lat, colour = whale_id),
            linewidth = 0.25, alpha = 0.6) +
  geom_point(data = data,
             aes(location.long, location.lat, colour = whale_id),
             size = 0.3, alpha = 0.8) +
  geom_point(data = data |> distinct(whale_id, .keep_all = TRUE),
             aes(location.long, location.lat),
             shape = 3, size = 3, colour = "red", stroke = 0.8) +
  scale_colour_viridis_d(guide = "none") +
  coord_sf(xlim = c(bbox_lims["left"],  bbox_lims["right"]),
           ylim = c(bbox_lims["bottom"], bbox_lims["top"])) +
  labs(x = "Longitude", y = "Latitude",
       title = "Bowhead whale telemetry tracks (all individuals)") +
  theme_minimal(base_size = 11)
```

## 3. Data preprocessing

```{r preprocessing}
# Remove low-quality Argos class Z, keep relevant columns
dataprep <- data |>
  filter(!(argos.lc %in% c("Z"))) |>
  select(ID = whale_id, timestamp,
         location.long, location.lat, argos.lc)

# Compute cumulative time in hours, remove duplicate timestamps
dataprep$time <- 0
rm_index <- logical(nrow(dataprep))

for (id in unique(dataprep$ID)) {
  idx   <- dataprep$ID == id
  steps <- as.numeric(diff(dataprep$timestamp[idx])) / 3600
  if (length(steps) > 0) {
    rm_index[idx] <- c(FALSE, steps == 0)
    dataprep$time[idx] <- c(0, cumsum(steps))
  }
}
dataprep <- dataprep[!rm_index, ]
```

## 4. Select focal individual and project coordinates

```{r focal-individual}
ID_FOCAL <- 128145

ggplot() +
  geom_sf(data = land, fill = "grey85", colour = "grey60",
          linewidth = 0.25) +
  geom_path(data = dataprep[dataprep$ID == ID_FOCAL, ],
            aes(location.long, location.lat),
            color = "steelblue", linewidth = 0.25, alpha = 0.6) +
  geom_point(data = dataprep[dataprep$ID == ID_FOCAL, ],
             aes(location.long, location.lat),
             size = 0.3, alpha = 0.8) +
  geom_point(data = dataprep[dataprep$ID == ID_FOCAL, ] |>
               distinct(ID, .keep_all = TRUE),
             aes(location.long, location.lat),
             shape = 3, size = 3, colour = "red", stroke = 0.8) +
  scale_colour_viridis_d(guide = "none") +
  coord_sf(xlim = c(bbox_lims["left"],  bbox_lims["right"]),
           ylim = c(bbox_lims["bottom"], bbox_lims["top"])) +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 11)
```

```{r project-coords}
data_SGD <- dataprep |>
  filter(ID == ID_FOCAL) |>
  rename(Y1 = location.long, Y2 = location.lat)

# Project to Canada Atlas Lambert (EPSG:3978), convert to km
data_SGD_sf   <- st_as_sf(data_SGD, coords = c("Y1", "Y2"), crs = 4326)
data_SGD_proj <- st_transform(data_SGD_sf, crs = 3978)
coords        <- st_coordinates(data_SGD_proj)
data_SGD$Y1   <- coords[, 1] / 1000
data_SGD$Y2   <- coords[, 2] / 1000

obs_ep <- make_argos_obs_params(dataprep[dataprep$ID == ID_FOCAL, ]$argos.lc)
```

## 5. Build the water-domain polygon

```{r water-polygon}
water_sf <- st_difference(bbox_poly, st_union(land)) |>
  st_collection_extract("POLYGON") |>
  st_make_valid() |>
  st_transform(crs = 3978)

polygon_coords <- do.call(rbind, lapply(
  seq_along(st_geometry(water_sf)), function(i) {
    geom <- st_geometry(water_sf)[[i]]
    do.call(rbind, lapply(seq_along(geom), function(j) {
      do.call(rbind, lapply(seq_along(geom[[j]]), function(k) {
        coords_k           <- as.data.frame(geom[[j]][[k]])
        colnames(coords_k) <- c("V1", "V2")
        coords_k$group     <- paste(i, j, k, sep = "_")
        coords_k
      }))
    }))
  }
))

polygon <- Polygon(1e-3 * as.matrix(
  polygon_coords[polygon_coords$group == "1_1_1", c("V1", "V2")]
))
```

```{r plot-focal-track}
poly_df        <- as.data.frame(polygon@coords)
single_track_df <- data_SGD

ggplot() +
  geom_polygon(data = poly_df,
               aes(x = V1, y = V2),
               fill = "#cce5f0", colour = "grey45", linewidth = 0.25) +
  geom_path(data = single_track_df,
            aes(x = Y1, y = Y2),
            colour = "darkorange", linewidth = 0.45, alpha = 0.8) +
  geom_point(data = single_track_df,
             aes(x = Y1, y = Y2),
             colour = "firebrick", size = 0.6, alpha = 0.5) +
  geom_point(data = single_track_df[1, ],
             aes(x = Y1, y = Y2),
             shape = 3, size = 4, colour = "black", stroke = 1) +
  labs(x = "Easting (km)", y = "Northing (km)",
       title = paste0("Bowhead whale ID ", ID_FOCAL, " — observed track")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## 6. Fit the smoothSDE baseline (Kalman filter)

```{r fit-kalman}
formulas <- list(mu1 = ~1, mu2 = ~1, tau = ~1, nu = ~1)

# Measurement error covariance from Brost et al. 2015
df_per_class    <- c(16.5, 2.3, 3, 3, 3, 3)
error_per_class <- sqrt(df_per_class / (df_per_class - 2)) *
  c(1.365, 1.089, 2.259, 2.720, 2.702, 13.338)
names(error_per_class) <- c("1", "2", "3", "0", "A", "B")

H <- array(rep(diag(2), nrow(dataprep)), dim = c(2, 2, nrow(dataprep)))
for (lc in c("0", "1", "2", "3", "A", "B")) {
  H[, , dataprep$argos.lc == lc] <- diag(rep(error_per_class[lc]^2, 2))
}

ctcrw <- SDE$new(formulas = formulas,
                 data     = data_SGD,
                 type     = "CTCRW",
                 response = c("Y1", "Y2"),
                 par0     = c(mu1 = 0, mu2 = 0, tau = 30, nu = 2),
                 other_data = list("H" = H[, , dataprep$ID == ID_FOCAL]),
                 fixpar   = c("mu1", "mu2"))
ctcrw$fit()

# Extract estimates and standard errors
sde_est <- as.list(ctcrw$tmb_rep(), what = "Estimate")
sde_se  <- as.list(ctcrw$tmb_rep(), what = "Std. Error")

log_tau_hat <- sde_est$coeff_fe[3]
log_nu_hat  <- sde_est$coeff_fe[4]
log_tau_se  <- sde_se$coeff_fe[3]
log_nu_se   <- sde_se$coeff_fe[4]

results_sde <- data.frame(
  Method   = "smoothSDE (Kalman filter)",
  param    = c("tau", "nu"),
  estimate = c(exp(log_tau_hat), exp(log_nu_hat)),
  se       = c(exp(log_tau_hat) * log_tau_se,
               exp(log_nu_hat)  * log_nu_se),
  ci_lower = c(exp(log_tau_hat - 1.96 * log_tau_se),
               exp(log_nu_hat  - 1.96 * log_nu_se)),
  ci_upper = c(exp(log_tau_hat + 1.96 * log_tau_se),
               exp(log_nu_hat  + 1.96 * log_nu_se)),
  row.names = NULL
)

print(results_sde)
```

## 7. Run particle filter with smoothSDE parameters

```{r particle-filter-setup}
sde_params <- list(tau   = ctcrw$par()[1, "tau"],
                   nu    = ctcrw$par()[1, "nu"],
                   omega = 0)

sigma_obs          <- 3
argos_error_params <- list(sigma_obs = sigma_obs, rho = 0.2,
                           a = 0.7, p = 0.5, df = 3)

x_star           <- matrix(c(680, 7050), ncol = 2, byrow = TRUE)
B_list           <- list(diag(c(1 / 50^2, 1 / 50^2)))
alpha            <- c(0)
potential_params <- list(x_star = x_star, B = B_list, alpha = alpha)
```

The particle filter below takes several minutes to run. The code is shown for
reproducibility; pre-computed results are loaded in the next chunk.

```{r particle-filter, eval=FALSE}
set.seed(2026)
filtered <- particle_filter2D_cpp(
  observations     = as.matrix(data_SGD[, c("time", "Y1", "Y2")]),
  sde_params       = sde_params,
  potential_params = potential_params,
  error_params     = argos_error_params,
  error_dist       = "argos",
  polygon_coords   = polygon@coords,
  U0               = as.numeric(c(data_SGD[1, c("Y1", "Y2")], 0, 0)),
  lambda           = sqrt(1.8),
  num_particles    = 1000,
  split_around_fixed_point = FALSE,
  scheme           = "Lie-Trotter",
  ESS_threshold    = 0.8,
  proposal_weight  = 0.5,
  verbose          = FALSE,
  print_timing     = FALSE,
  obs_error_params = obs_ep
)

saveRDS(filtered, system.file("extdata", "penalized_filtered_bowhead.rds",
                              package = "LangevinAnimalTracking"))
```

```{r particle-filter-load, echo=FALSE}
filtered <- readRDS(system.file("extdata", "penalized_filtered_bowhead.rds",
                                package = "LangevinAnimalTracking"))
```

```{r particle-filter-estimates}
estimates <- as.data.frame(get_PF_mean(filtered))
```

## 8. Run Kalman filter (no penalty)

```{r kalman-no-penalty}
filtered_nopenalty <- kalman_filter(
  data_SGD[, c("time", "Y1", "Y2")],
  sde_params,
  potential_params = potential_params,
  sigma_obs        = H[, , dataprep$ID == ID_FOCAL],
  polygon,
  lambda           = Inf,
  U0               = as.numeric(c(data_SGD[1, c("Y1", "Y2")], 0, 0)),
  R0               = diag(c(1, 1, 1, 1)),
  split_around_fixed_point = FALSE
)

estimates_nopenalty <- as.data.frame(filtered_nopenalty$U) |>
  rename(X1 = V1, X2 = V2, V1 = V3, V2 = V4)
```

## 9. Compare fraction of locations on land

```{r fraction-on-land}
polygon_sf <- st_sfc(st_polygon(list(polygon@coords)), crs = 32616)

filtered_sf_pen   <- st_as_sf(estimates,         coords = c("X1", "X2"), crs = 32616)
filtered_sf_nopen <- st_as_sf(estimates_nopenalty, coords = c("X1", "X2"), crs = 32616)

pct_land_pen   <- mean(!st_within(filtered_sf_pen,   polygon_sf, sparse = FALSE))
pct_land_nopen <- mean(!st_within(filtered_sf_nopen, polygon_sf, sparse = FALSE))

cat("Fraction on land - with penalization:    ", round(pct_land_pen,   4) * 100, "\n")
cat("Fraction on land - without penalization: ", round(pct_land_nopen, 4) * 100, "\n")
```

## 10. Compare tracks from both methods

```{r plot-comparison}
estimates$method           <- "Penalized particle filter"
estimates_nopenalty$method <- "Kalman filter"
both <- rbind(estimates, estimates_nopenalty)

ggplot() +
  geom_polygon(data = poly_df,
               aes(x = V1, y = V2),
               fill = "#cce5f0", colour = "grey45", linewidth = 0.25) +
  geom_point(data = data_SGD,
             aes(x = Y1, y = Y2),
             colour = "red", size = 0.4, alpha = 0.4) +
  geom_path(data = both,
            aes(x = X1, y = X2, colour = method),
            linewidth = 0.5, alpha = 0.8) +
  scale_colour_manual(values = c("Penalized particle filter" = "darkorange",
                                 "Kalman filter"             = "steelblue")) +
  labs(x = "Easting (km)", y = "Northing (km)", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

```{r plot-comparison-zoom}
# Zoom on a region where differences are most visible
ggplot() +
  geom_polygon(data = poly_df,
               aes(x = V1, y = V2),
               fill = "#cce5f0", colour = "grey45", linewidth = 0.25) +
  geom_point(data = data_SGD,
             aes(x = Y1, y = Y2),
             colour = "red", size = 0.4, alpha = 0.4) +
  geom_path(data = both,
            aes(x = X1, y = X2, colour = method),
            linewidth = 0.5, alpha = 0.8) +
  scale_colour_manual(values = c("Penalized particle filter" = "darkorange",
                                 "Kalman filter"             = "steelblue")) +
  coord_cartesian(xlim = c(500, 700), ylim = c(1800, 2500)) +
  labs(x = "Easting (km)", y = "Northing (km)", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

## 11. Estimate parameters with SGD

The SGD optimisation requires approximately 7 hours on a standard laptop.
The code below shows the exact call used to produce the pre-saved results
that are loaded in the next chunk.

```{r sgd-fit, eval=FALSE}
theta_init <- list(tau = 30, nu = 2, omega = 0)

set.seed(2026)
sgd <- SGD_Fisher(
  data             = data_SGD,
  sde_params       = theta_init,
  fixpar           = c("omega"),
  SGD_iter         = 200,
  potential_params = potential_params,
  error_dist       = "argos",
  error_params     = argos_error_params,
  scheme           = "Lie-Trotter",
  polygon          = polygon,
  U0               = as.numeric(c(data_SGD[1, c("Y1", "Y2")], 0, 0)),
  lambda           = sqrt(1.8),
  num_particles    = 500,
  split_around_fixed_point = FALSE,
  verbose          = TRUE,
  gamma0           = 1e-4,
  K_preheat        = 50,
  alpha            = 2/3,
  C_heating        = 1/10,
  n_smooth_samples = 10,
  obs_error_params = obs_ep
)

saveRDS(sgd, system.file("extdata", "sgd_bowhead.rds",
                         package = "LangevinAnimalTracking"))
```

```{r sgd-load, echo=FALSE}
sgd <- readRDS(system.file("extdata", "sgd_bowhead.rds",
                           package = "LangevinAnimalTracking"))
```

## 12. Extract SGD estimates and confidence intervals

```{r sgd-results}
Delta_final <- sgd$Delta
I_fisher    <- t(Delta_final) %*% Delta_final
Sigma_sgd   <- solve(I_fisher)
se_sgd      <- sqrt(diag(Sigma_sgd))

theta_hat <- sgd$theta[nrow(sgd$theta), ]

results_sgd <- data.frame(
  Method   = "SGD (Lie-Trotter)",
  param    = names(theta_hat),
  estimate = theta_hat,
  se       = se_sgd,
  ci_lower = theta_hat - 1.96 * se_sgd,
  ci_upper = theta_hat + 1.96 * se_sgd,
  row.names = NULL
)

print(results_sgd)
```

## 13. Comparison table: SGD vs Kalman filter

```{r comparison-table}
results_sgd_sub <- results_sgd |>
  filter(param %in% c("tau", "nu")) |>
  select(Method, param, estimate, ci_lower, ci_upper)

results_sde_sub <- results_sde |>
  select(Method, param, estimate, ci_lower, ci_upper)

comparison_table <- bind_rows(results_sgd_sub, results_sde_sub) |>
  arrange(param, Method)

print(comparison_table)
```

## 14. Posterior track reconstruction

```{r plot-posterior}
posterior_df <- do.call(rbind, lapply(
  seq_along(sgd$smooth_samples), function(i) {
    traj        <- as.data.frame(sgd$smooth_samples[[i]][, c("X1", "X2")])
    traj$sample <- i
    traj
  }
))

ggplot() +
  geom_polygon(data = poly_df,
               aes(x = V1, y = V2),
               fill = "#cce5f0", colour = "grey45", linewidth = 0.25) +
  geom_path(data = posterior_df,
            aes(x = X1, y = X2, group = sample),
            colour = "darkorange", alpha = 0.45, linewidth = 0.35) +
  geom_point(data = data_SGD,
             aes(x = Y1, y = Y2),
             colour = "firebrick", size = 0.7, alpha = 0.55) +
  labs(x = "Easting (km)", y = "Northing (km)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## 15. SGD convergence check

```{r plot-convergence}
theta_SGD           <- data.frame(sgd$theta)
theta_SGD$iteration <- seq_len(nrow(theta_SGD))

df_long <- theta_SGD |>
  pivot_longer(cols = c("tau", "nu"),
               names_to = "parameter", values_to = "value")

convergence_theme <- theme_minimal(base_size = 14) +
  theme(
    strip.text         = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "none"
  )

ggplot(df_long, aes(x = iteration, y = value, color = parameter)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 1,
             labeller = label_parsed) +
  scale_color_manual(values = c(tau = "steelblue", nu = "darkorange")) +
  labs(x = "Iteration", y = "Value") +
  convergence_theme
```
